Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-10-09 paper.tex
.

@@

\comment[CC]{Juste, tu penses qu'il faudrait faire la meme analyse que toi sur la distance entre phase diagrams? Ce serait cool mais si oui, j'aurais besoin d'aide (et des fichiers de simu!)}

\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup \comment[CC]{j'ai peut-etre manque une etape, mais c'est quoi le default model setup? en termes de grille d'entree: uniforme? random? ou Netlogo?}[(JR) c'est Netlogo, une configuration bi-centrique particulière] (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability. \comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each. Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process). \comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Red point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure}
%%%%%%%%%%%%%

@@

\subsection{Qualitative variations}
\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams. We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value. These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}. The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories. The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values. The process is thus completely inverted, what would have significant impacts if one tried to schematize policies from this model. This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.

\comment{Ajouter un paragraphe sur le lien entre forme de grille et comportement du modèle/résultats.}

\paragraph{Schelling} In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf. figure \ref{fig:densityTypes}). In particular, we want evaluate to which extent the typology summarizes the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters. This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.

>>>>

We chose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method.

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams}
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} \label{fig:schelling-distance-meta}}
\end{figure}
%%%%%%%%%%%%%

\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability. \comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each. Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Black point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure}
%%%%%%%%%%%%%

@@

\subsection{Qualitative variations}
\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams. We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value. These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}. The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories. The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values. The process is thus completely inverted, what would have significant impacts if one tried to schematise policies from this model. To interpret these behavior in terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second. In sprawled spaces, inequalities are thus fostered by a lack of minimal local ressources, whereas population will drive inequality in concentrated spaces.

%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.

\paragraph{Schelling} In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf. figure \ref{fig:densityTypes}). In particular, we want evaluate to which extent the typology summarises the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters. This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
